Exp 13 write-up

### 2023-08-17 writing/DGfS2023-abstract/abstract.tex
updated DGfS abstract

\definecolor{airforceblue}{rgb}{0.36, 0.54, 0.66}
%\definecolor{gray}{rgb}{0.36, 0.54, 0.66}

\definecolor{Pink}{RGB}{240,0,120}
\newcommand{\red}[1]{\textcolor{Pink}{#1}}
\newcommand{\jd}[1]{\textbf{\textcolor{Pink}{[jd: #1]}}}

@@

\citealt{mandelkern-etal2020} challenged this conclusion, arguing that gradient projection is an artefact of the inference task used, writing that ``inference tasks to some degree invite subjects to make the inference in question'' (p.498) and that ``we should think twice before embracing a notion of presupposition projection that is gradient based on results from inference tasks alone'' (p.497). They suggested that naturalness ratings of utterances with attitude predicates in explicit ignorance contexts (EICs), as in (\ref{second}), are more suitable to distinguish the semantic presuppositions of factive predicates (hypothesized to be unnatural in EICs) from inferences that arise with nonfactive predicates ``for any of a variety of pragmatic reasons short of entailment or presupposition'' (p.497).

@@

\noindent
\underline{Participants.} Data from 370 participants (ages: 19-80, mean: 40.7; 175 women, 185 men, 8 nonbinary, 2 undisclosed) were analyzed. Other data were excluded based on preregistered criteria.

\noindent
\underline{Materials.} Participants read two-sentence discourses consisting of a declarative followed by an interrogative, like (\ref{second}). In the target stimuli, the interrogatives combined the 20 factive and nonfactive predicates with the 20 complement clauses of \citealt{degen-tonhauser-openmind,degen-tonhauser-language}. The preceding declaratives implemented a three-level context condition: In the `explicit ignorance' context (\ref{second}), the declarative sentence conveyed the speaker's ignorance about the CC: If the CC is presupposed, global accommodation (taken to be default in \citealt{heim83,vds92}) is not possible and the discourse is expected to be unnatural; if the CC is not presupposed, the discourse is expected to be natural. In the other two context levels, the declarative denoted a fact relative to which the CC had a `low prior probability' (\ref{sample}a) or a `high prior probability' (\ref{sample}b); these contexts were normed in \citealt{degen-tonhauser-openmind}. Presupposition analyses (e.g., \citealt{heim83,vds92}) invariably predict global accommodation for presupposed CCs, that is, high naturalness ratings.

@@

In addition to the 1,200 target stimuli (20 predicates $\times$ 20 CCs $\times$ 3 contexts), the materials included six fillers: Four with hard triggers ({\em too, also, again},  {\em it-}clefts), expected to be unnatural in EICs (\citealt{simons01,abusch10}), and two with {\em stop} and {\em continue}, which \citealt{simons01} considers soft triggers (natural in explicit ignorance contexts), contrary to \citealt{mandelkern-etal2020} and \citealt{kalomoiros-schwarz2021}. The fillers were only presented in EICs.

@@

\noindent
{\bf Results.} Contrary to \citepos{mandelkern-etal2020} claim, factives do not pattern alike, as shown in Fig.~\ref{fig:acc-by-expression}, which plots mean naturalness ratings in EICs by expression (colors: \color{orange}factives\color{black}, \color{gray}nonfactives\color{black}, hard triggers, \color{magenta}soft triggers\color{black}). Whereas the mean for {\em be annoyed} is at the level of the soft trigger {\em stop}, that of {\em know} is higher, and the means of {\em see, discover}, and {\em reveal} are as high as those of some nonfactive predicates. (These observations are confirmed by a linear mixed effects model, shown in the talk.) Thus, in line with \citealt{degen-tonhauser-language}, there is no empirical evidence for a categorical distinction between factive and non-factive predicates. (The talk also discusses that the results do not align with assumptions about hard and soft triggers; \citealt{simons01,abusch10}.)

Contrary to the predictions of presupposition analyses, the naturalness ratings of the purportedly factive predicates are sensitive to the low vs.\ high prior probability context manipulation. As shown in Fig.~\ref{fig:acc-by-context}, which plots mean naturalness ratings by context and expression, the mean naturalness ratings are higher in the high prior probability context than in the low one. This context-sensitivity, which is also observed for the nonfactive predicates (except {\em pretend}; model details provided in talk), replicates \citepos{degen-tonhauser-openmind} results, obtained using an inference rating task.

\noindent
{\bf Methodological and theoretical implications.} The recent challenge to the categorical factivity distinction by \citealt{degen-tonhauser-language} is not addressed by collecting naturalness ratings in EICs instead of inference ratings. And while naturalness ratings in EICs may provide insight into the meanings of attitude predicates, the interpretation of the results depends on the presumed linking function (as discussed in the talk; see, e.g., \citealt{sprouse2018}). The very low mean naturalness rating for {\em be annoyed} in EICs could, for instance, be due not to a semantic presupposition (see also \citealt{karttunen2016}) but because the CC is not-at-issue, contrary to what is suggested by the EIC.

Contemporary presupposition analyses (e.g., \citealt{heim83,vds92,abrusan2011,karttunen2016,simons-etal2017}) predict neither the heterogeneity of the naturalness ratings of factive predicates nor their sensitivity to the prior probability of the CC. These results, which mirror those based on inference ratings, suggest that projection analyses must incorporate more fine-grained distinctions between attitude predicates and the systematic effect of prior probabilities.

@@

\begin{figure}[h!]
\centering
\caption{\small{Mean naturalness rating in explicit ignorance context by expression (\color{orange}factive\color{black}, \color{gray}nonfactive\color{black}, \color{magenta} soft trigger\color{black}, \color{black}hard trigger\color{black}). Error bars indicate 95\% bootstrapped CIs. Violin plots indicate kernel probability density of participants' ratings.}}\label{fig:acc-by-expression}
\includegraphics[width=.9\textwidth]{../../results/main/13explicitIgnorance/graphs/explicit-ignorance-naturalness-by-predicate}
\end{figure}

>>>>

\definecolor{green}{RGB}{0,158,115}
\definecolor{orange}{RGB}{213,94,0}

\newcommand{\red}[1]{\textcolor{Pink}{#1}}
\newcommand{\jd}[1]{\textbf{\textcolor{Pink}{[jd: #1]}}}

@@

\citealt{mandelkern-etal2020} challenged this conclusion, arguing that gradient projection is an artefact of the inference task used, writing that ``inference tasks to some degree invite subjects to make the inference in question'' (p.498) and that ``we should think twice before embracing a notion of presupposition projection that is gradient based on results from inference tasks alone'' (p.497). They suggested that naturalness ratings of utterances with attitude predicates in explicit ignorance contexts (EICs), as in (\ref{second}), are more suitable to distinguish the semantic presuppositions of factive predicates (hypothesized to be unnatural in EICs) from inferences that may arise with nonfactive predicates ``for any of a variety of pragmatic reasons short of entailment or presupposition'' (p.497).

@@

\noindent
\underline{Participants.} 398 participants were recruited on Prolific. The data of 370 entered into the analysis.

\noindent
\underline{Materials.} Participants read two-sentence discourses consisting of a declarative followed by an interrogative, like (\ref{second}). In the target stimuli, the interrogatives combined the 20 (non)factive predicates with the 20 complement clauses of \citealt{degen-tonhauser-openmind,degen-tonhauser-language}. The preceding declaratives implemented a three-level context condition: In the `explicit ignorance' context (\ref{second}), the declarative sentence conveyed the speaker's ignorance about the CC: If the CC is presupposed, global accommodation (taken to be default in \citealt{heim83,vds92}) is not possible and the discourse is expected to be unnatural; if the CC is not presupposed, the discourse is expected to be natural. In the other two context levels, the declarative denoted a fact relative to which the CC had a `low prior probability' (\ref{sample}a) or a `high prior probability' (\ref{sample}b); these contexts were normed in \citealt{degen-tonhauser-openmind}. Presupposition analyses (e.g., \citealt{heim83,vds92}) invariably predict global accommodation for presupposed CCs, that is, high naturalness ratings.

@@

In addition to the 1,200 target stimuli (20 predicates $\times$ 20 CCs $\times$ 3 contexts), the materials included six fillers: Four with hard triggers ({\em too, also, again},  {\em it-}clefts), expected to be unnatural in EICs (\citealt{simons01,abusch10}), and two with {\em stop} and {\em continue}, which \citealt{simons01} considers soft triggers (natural in EICs), contrary to \citealt{mandelkern-etal2020} and \citealt{kalomoiros-schwarz2021}, who take them to be unnatural in EICs. The fillers were only presented in EICs.

@@

\noindent
{\bf Results.} Contrary to \citepos{mandelkern-etal2020} claim, factives do not pattern alike, as shown in Fig.~\ref{fig:acc-by-expression}, which shows mean naturalness ratings in EICs by expression (colors: \color{orange}factives\color{black}, \color{gray}nonfactives\color{black}, hard triggers, \color{green}soft triggers\color{black}). Whereas the mean for {\em be annoyed} is at the level of the soft trigger {\em stop}, that of {\em know} is higher, and the means of {\em see, discover}, and {\em reveal} are as high as those of some nonfactive predicates. (These observations are confirmed by a linear mixed effects model.) Thus, in line with \citealt{degen-tonhauser-language}, there is no empirical evidence for a categorical distinction between factive and non-factive predicates. (The talk also discusses that the results do not align with assumptions about hard and soft triggers; \citealt{simons01,abusch10}.)

Contrary to the predictions of presupposition analyses, the naturalness ratings of the purportedly factive predicates are sensitive to the low vs.\ high prior probability context manipulation. As shown in Fig.~\ref{fig:acc-by-context}, which shows mean naturalness ratings by context and expression, the mean naturalness ratings are higher in the high prior probability context than in the low one. This context-sensitivity, which is also observed for the nonfactive predicates (except {\em pretend}), replicates \citepos{degen-tonhauser-openmind} results, obtained using an inference rating task.

\noindent
{\bf Methodological and theoretical implications.} The recent challenge to the categorical factivity distinction by \citealt{degen-tonhauser-language} is not dispelled by collecting naturalness ratings in EICs instead of inference ratings. And while naturalness ratings in EICs may provide insight into the meanings of attitude predicates, the interpretation of the results depends on the presumed linking function (see, e.g., \citealt{sprouse2018}). Instead of the ratings directly reflecting whether the CC is presupposed, the very low mean rating for {\em be annoyed} in EICs could be due to its CC being not-at-issue while the EIC suggests that it is at-issue; likewise, reading {\em know} with a contrastive pitch accent in EICs may have resulted in about half of the participants giving a high rating.

Contemporary presupposition analyses (e.g., \citealt{heim83,vds92,abrusan2011,karttunen2016,simons-etal2017}) predict neither the heterogeneity of the naturalness ratings of factive predicates nor their sensitivity to the prior probability of the CC (though see \citealt{schlenker2021} for the latter). These results, which mirror those based on inference ratings, suggest that projection analyses must incorporate more fine-grained distinctions between attitude predicates and the systematic effect of prior probabilities on projection.

@@

\begin{figure}[h!]
\centering
\caption{\small{Mean naturalness rating in explicit ignorance context by expression (\color{orange}factive\color{black}, \color{gray}nonfactive\color{black}, \color{green} soft trigger\color{black}, \color{black}hard trigger\color{black}). Error bars indicate 95\% bootstrapped CIs. Violin plots indicate kernel probability density of participants' ratings.}}\label{fig:acc-by-expression}
\includegraphics[width=.9\textwidth]{../../results/main/13explicitIgnorance/graphs/explicit-ignorance-naturalness-by-predicate}
\end{figure}
